---
date: "Date: `r Sys.Date()`"
output:
  html_document:
      toc: true
      toc_depth: 3
      toc_float: true
      theme: cerulean
geometry: margin = 2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


required.libraries <- c("cowplot",
                        "data.table",
                        "IRanges",
                        "rmarkdown",
                        "tidyverse")

for (lib in required.libraries) {
  suppressPackageStartupMessages(library(lib, character.only = TRUE, quietly = T))
}

# Load utilities
source(file = "Utilities.R")

# Load table with protease clavage results
proteasome.cleavage.results <- fread("data/Proteasome_cleavage_results.txt") |> 
                                within(rm(cleaved))
```


## Define a S4 class

Lets define an `S4` class containing the files needed for the analysis of protein cleavage sites.

```{r ProteasomeCleavage, echo = TRUE}
# --------------- #
# Define S4 class #
# --------------- #

# This is the definition of the ProteasomeCleavage class (S4 object)
#
# 1) slots: define the slots (a.k.a. fields) included in the new class, each slot with its own class
#
# 2) prototype: default values, NA otherwise.
#
# The names in 'slots' and 'prototype' must be the same, otherwise, there will
# be an error.

setClass("ProteasomeCleavage",
         
         slots     = c(id                 = "character",
                       sequence           = "character",
                       sequence_length    = "numeric",
                       linker             = "character",
                       nb_linkers         = "numeric",
                       cleavage_threshold = "numeric",
                       nb_sites           = "numeric",
                       cleavage_score     = "numeric",
                       fragments_imm      = "data.table"),
         
         prototype = list(id                 = NA_character_,
                          sequence           = NA_character_,
                          sequence_length    = NA_real_,
                          linker             = NA_character_,
                          nb_linkers         = NA_real_,
                          cleavage_threshold = 0.5,
                          nb_sites           = NA_real_,
                          cleavage_score     = NA_real_,
                          fragments_imm      = data.table()),
         
         validity = check_ProteasomeCleavage
)
```

<p>&nbsp;</p>
<p>&nbsp;</p>

## Validity function

- Define a *validity* function that checks all the arguments have the correct format (e.g., 
data.frames with a specified number of columns, the lengt of some arguments that must be strictly of length 1, etc.) and returns the error messages, if any.

- This function will be ran each time a `S4` instance is modified.

- This *validity* function must have only one argument `object`.

- Output: `TRUE` if all the validty test were passed, or all the collected error messages. 

```{r Validity_example, echo = TRUE} 
# Use this function to verify the data format is correct
# Ideally, you want to verify each slot
check_ProteasomeCleavage <- function(object) {
  
  # Collect error messages in this vector
  errors <- character()
  
  # ID
  length.id <- length(object@id)
  if (length.id != 1) {
    msg <- paste("ID is length ", length.id, ".  Should be 1", sep = "")
    errors <- c(errors, msg)
  }
  
  
  # Sequence
  if (!is.na(object@sequence)) {
    
    nchar.seq <- nchar(object@sequence)
    
    if (nchar.seq < 1) {
      msg <- paste("Sequence length ", nchar.seq, ".  Should be a positive integer", sep = "")
      errors <- c(errors, msg)
    }
  }
  
  
  # Sequence length
  length.seq <- length(object@sequence_length)
  if (length.seq != 1) {
    msg <- paste("The number of sequence lengths is ", length.seq, ".  Should be 1", sep = "")
    errors <- c(errors, msg)
  }
  
  
  # Linker
  if (!is.na(object@linker)) {
    
    length.linker <- length(object@linker)
    
    if (length.linker != 1) {
      msg <- paste("The number of linkers is ", length.linker, ".  Should be 1", sep = "")
      errors <- c(errors, msg)
    }
  }
  
  
  # Number of Linkers
  if (!is.na(object@nb_linkers)) {
    
    length.nb_linkers <- length(object@nb_linkers)
    
    if (length.nb_linkers != 1) {
      msg <- paste("The number of linker lengths is ", length.linker, ".  Should be 1", sep = "")
      errors <- c(errors, msg)
    }
  }
  
  # Treshold
  if (!is.na(object@cleavage_threshold)) {
    
    length.cleavage_threshold <- length(object@cleavage_threshold)
    
    if (length.cleavage_threshold != 1) {
      msg <- paste("The number of thresholds is ", length.seq, ".  Should be 1", sep = "")
      errors <- c(errors, msg)
    }
    
    if (object@cleavage_threshold < 0 | object@cleavage_threshold > 1){
      msg <- paste("The threshold value is ", object@cleavage_threshold, ".  Should be between 0-1", sep = "")
      errors <- c(errors, msg)
    }
  }
  
  
  # Number of sites
  if (!is.na(object@nb_sites)) {
    
    length.nb_sites <- length(object@nb_sites)
    
    if (length.nb_sites != 1) {
      msg <- paste("The number of sites length is ", length.nb_sites, ".  Should be 1", sep = "")
      errors <- c(errors, msg)
    }
    
    if (object@nb_sites < 1){
      msg <- paste("The number of sites value is ", object@nb_sites, ".  Should be a positive integer.", sep = "")
      errors <- c(errors, msg)
    }
    
    if (!is.integer(object@nb_sites)) {
      msg <- paste("The number of sites value is of class ", class(object@nb_sites), ".  Should be a positive integer.", sep = "")
      errors <- c(errors, msg)
    }
  }
  
  
  # Cleavage score
  if (!length(object@cleavage_score)) {
    
    length.cleavage_score <- length(object@cleavage_score)
    
    if (length.cleavage_score < 1){
      msg <- paste("The number of cleavage scores is ", length.cleavage_score, ".  Should be a positive integer", sep = "")
      errors <- c(errors, msg)
    }
  }
  
  
  # Fragments
  if (nrow(object@fragments_imm)) {
    
    ncol.fragments_imm <- ncol(object@fragments_imm)
    nrow.fragments_imm <- nrow(object@fragments_imm)
    
    if (nrow.fragments_imm < 1){
      msg <- paste("The number of rows in the fragments table is ", nrow.fragments_imm, ".  Should be a positive integer", sep = "")
      errors <- c(errors, msg)
    }
    
    if (ncol.fragments_imm != 6) {
      msg <- paste("The number of rows in the fragments table is ", ncol.fragments_imm, ".  Should be 6", sep = "")
      errors <- c(errors, msg)
    }
  }
  
  # In case of errors, return the collected messages
  if (length(errors) == 0) TRUE else errors
}
```

<p>&nbsp;</p>
<p>&nbsp;</p>

## Manual creation of an instance of the `ProteasomeCleavage` class

<p>&nbsp;</p>

- Use the function `new` to create an empty instance of a `S4` class object.

```{r ProteasomeCleavage_manual_creation, echo = TRUE}

# This is an initialization, the remaining fields will be filled up afterwards
ID1.ProteasomeCleavage <- new("ProteasomeCleavage",
                              id       = "ID1",
                              sequence = "CAHVIWGFDMDVYSMDGTSMFYILNFS")

# Uncomment for example only
# ID1.ProteasomeCleavage <- new("ProteasomeCleavage",
#                               id  = "ID1",
#                               seq = "CAHVIWGFDMDVYSMDGTSMFYILNFS")

```

<p>&nbsp;</p>

- Use the function `getSlots` to show the class prototype.

- Show the content of the new instance. Note that the slot `cleavage_threshold` was already
initialize from the prototype when `ProteasomeCleavage` was defined.

```{r ProteasomeCleavage_manual_creation_slots, echo = TRUE, collapse=TRUE}

getSlots("ProteasomeCleavage")


ID1.ProteasomeCleavage

```

<p>&nbsp;</p>

- Use the operator `@` to acces the slots of a `S4` class object.
- Alternatively, you can use the function `slot`, with the instance of your class and the slot name.

```{r ProteasomeCleavage_access_slots, echo = TRUE, collapse=TRUE}

ID1.ProteasomeCleavage@id


slot(ID1.ProteasomeCleavage, "id")

```

<p>&nbsp;</p>
<p>&nbsp;</p>

## Automated creation of an instance of the `ProteasomeCleavage` class

- Let's explore how many sequences we have in our results and their length.
- In this example we will create one instance for each protein ID in the `proteasome.cleavage.results` table.

```{r Explore_results, echo = TRUE, collapse=TRUE}

# Number of protein IDs in our result table
unique(proteasome.cleavage.results$protein_id)


# Sequence length of each protein
proteasome.cleavage.results |> 
  group_by(protein_id) |> 
  summarise(Nb_residues = n())
```

<p>&nbsp;</p>
<p>&nbsp;</p>

- The code of the function to generate the instances is in the `Utilities.R` script in this repository.

```{r ProteasomeCleavage_automated_creation, echo = TRUE, collapse=TRUE}

# ProteasomeCleavageFeatures is a list where each entry contains a ProteasomeCleavage object
ProteasomeCleavageFeatures <- CreateProteasomeCleavageFeatures(in.df = proteasome.cleavage.results)


names(ProteasomeCleavageFeatures)


ProteasomeCleavageFeatures$Sequence_1
```

<p>&nbsp;</p>
<p>&nbsp;</p>

## Update slots of a `ProteasomeCleavage` instance.

- Some slots can be updated once the `ProteasomeCleavage` instance was created.
- We will add the linker sequence to each protein in our example.

```{r ProteasomeCleavage_update_slot, echo = TRUE, collapse=TRUE}

ProteasomeCleavageFeatures$Sequence_1@linker <- ""
ProteasomeCleavageFeatures$Sequence_2@linker <- "GGGGSGGGGS"
ProteasomeCleavageFeatures$Sequence_3@linker <- "GGGGSGGGGSGGGGS"

# Trying to assign a value not defined in the prototype will raise an error
# ProteasomeCleavageFeatures$Sequence_1@linker <- NA
# ProteasomeCleavageFeatures$Sequence_1@linker <- 1

ProteasomeCleavageFeatures$Sequence_3
```


<p>&nbsp;</p>
<p>&nbsp;</p>

## Define and apply methods taking advantage of our custom class

We have to complete the missing slots in our `ProteasomeCleavage` instances: `nb_linkers`, 
`nb_sites`, and `fragments_imm`.

One possibility is to calculate these values externally and then update the slots,
but we can also create methods that work with the `ProteasomeCleavage`.

<p>&nbsp;</p>

- We will find the number of linkers in the input sequences.
- First externally and after defining a method.

```{r Linkers_no_method, echo = TRUE, collapse=TRUE}

linker   <- ProteasomeCleavageFeatures$Sequence_2@linker
sequence <- ProteasomeCleavageFeatures$Sequence_2@sequence

# Then we can assign it to the ProteasomeCleavage instance in the 
ProteasomeCleavageFeatures$Sequence_2@nb_linkers <- str_count(sequence, pattern = linker)
```

<p>&nbsp;</p>

- We can also define a method that recognize the structure of our class.
- To define a method we need two steps:
    - Define a generic : let the system know the method name and the required arguments
    - Define the method : the function itself
- **NOTE**: the name and arguments provided to the `setGeneric` and `setMethod` must be the same.

```{r set_Linkers_method, echo = TRUE, collapse=TRUE}

# Define the generic
# 
# Requires two arguments:
# 1) The generic/method name
# 2) The method arguments
setGeneric("set_nb_linkers", function(x, value) standardGeneric("set_nb_linkers"))


# Define the method
#
# Requires three arguments:
#
# 1) Method name: Same name as in its corresponding 'setGeneric'
# 2) Class name: S4 class name
# 3) Arguments: similar when defining a function 
setMethod("set_nb_linkers", "ProteasomeCleavage", function(x, value) {
  x@nb_linkers <- value
  x 
  # Returns the updated version of the instance
})



ProteasomeCleavageFeatures$Sequence_3 <- set_nb_linkers(x     = ProteasomeCleavageFeatures$Sequence_3,
                                                        value = 7)


# These are not the real number of linkers, it is just an example
nb.linkers <- c(3,5,7)

ProteasomeCleavageFeatures <- map2(.x = ProteasomeCleavageFeatures,
                                   .y = nb.linkers,
                                   .f = ~set_nb_linkers(x     = .x,
                                                        value = .y))
```

<p>&nbsp;</p>
<p>&nbsp;</p>

- Now that we have understood the logic, let's create another method to calculate the number of linkers.
- This method calls the previosly defined `set_nb_linkers` method.

```{r Linkers_with_method, echo = TRUE, collapse=TRUE}

setGeneric("count_linkers_in_seq", function(x) standardGeneric("count_linkers_in_seq"))

setMethod("count_linkers_in_seq", "ProteasomeCleavage", function(x) {

  # First claculate the number of linkers ...
  if (x@linker != "") {
    
    # The sequence and linkers are taken from the CleavageFeatures instance ('x')
    nb.linkers <- str_count(x@sequence, pattern = x@linker)

    # If no linker sequence is provided, then the number of linkers becomes 0 
  } else {
    nb.linkers <- 0
  }
  
  # ... then write it
  x <- set_nb_linkers(x     = x,
                      value = nb.linkers)

  x
})


# Calculate the number of linkers
ProteasomeCleavageFeatures <- map(ProteasomeCleavageFeatures, count_linkers_in_seq)


ProteasomeCleavageFeatures$Sequence_1@nb_linkers


ProteasomeCleavageFeatures$Sequence_2@nb_linkers


ProteasomeCleavageFeatures$Sequence_3@nb_linkers
```

<p>&nbsp;</p>
<p>&nbsp;</p>

- Here is another little more complex example.

```{r nb_cleavage_sites, echo = TRUE, collapse=TRUE}
setGeneric("set_nb_sites", function(x, value) standardGeneric("set_nb_sites"))
setGeneric("set_fragments_imm", function(x, value) standardGeneric("set_fragments_imm"))
setGeneric("get_fragments", function(x) standardGeneric("get_fragments"))


setMethod("set_nb_sites", "ProteasomeCleavage", function(x, value) {
  x@nb_sites <- value
  x 
})


setMethod("set_fragments_imm", "ProteasomeCleavage", function(x, value) {
  x@fragments_imm <- value
  x 
})


setMethod("get_fragments", "ProteasomeCleavage", function(x) {

  # Get the number of sites with a score equal/greater than the threshold
  sites.pos <- which(x@cleavage_score >= x@cleavage_threshold)
  nb.sites  <- length(sites.pos) 
  
  x <- set_nb_sites(x     = x,
                    value = nb.sites)
  
  
  # Get the position of the fragments
  fragments.coordinates.df <- get.fragments.coordinates(cleavage.scores = x@cleavage_score,
                                                        thr             = x@cleavage_threshold,
                                                        proteinseq      = x@sequence,
                                                        name            = x@id)

  x <- set_fragments_imm(x     = x,
                         value = fragments.coordinates.df)
  x
})



# Get the fragment sequences
ProteasomeCleavageFeatures <- map(ProteasomeCleavageFeatures, get_fragments)


ProteasomeCleavageFeatures$Sequence_1@fragments_imm


ProteasomeCleavageFeatures$Sequence_2@fragments_imm


ProteasomeCleavageFeatures$Sequence_3@fragments_imm
```