---
title: "Retrieving and analyzing indegrees from cell-specific networks"
subtitle: "Indegree clustering" 
author:
    - Ine Bonthuis
date: 'Date: `r Sys.Date()`'
output:
    html_document:
        toc: yes
        toc_depth: 3
        toc_float: yes
        theme: cerulean
        code_download: true
    pdf_document:
        toc: yes
        toc_depth: '3'
geometry: margin = 1cm
---

```{r load_packages}
library(pbapply)
library(irlba)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(parallel)
library(tidyverse)
library(plyr)
library(dplyr)
library(tidyr)
require(Metrics)
library(data.table)
```

```{r standard_plotting_settings, eval = TRUE, echo = FALSE, include = TRUE, cache = TRUE}

standard_colors <- RColorBrewer::brewer.pal(8, "Set1")
umap_pal <- brewer.pal(7, "YlGnBu")

ggplot2::theme_set(
    theme_bw() +
        theme(
            plot.title = element_text(size = 14),
            axis.title = element_text(size = 14, colour = "black"),
            axis.text = element_text(size = 12, colour = "black")
        )
)

```


For the following umap plot, I'd like to distinguish the cell types based on both color and shapes, the shapes are useful because it allows for visualization of overlaps. But I didn't manage to get that into legend yet, so how can I put this in the legend?

```{r umap}
umapdata <- fread( "ine/umapdata.tsv")
umapdata <- as.data.frame(umapdata)
rownames(umapdata) <- umapdata$V1 
umapdata$V1 <- NULL
head(umapdata)

umapdata$celltypes <- as.factor(umapdata$celltypes)

celltypes <- umapdata$celltypes

umapPlot <- ggplot(umapdata, aes(X1, X2, color = celltypes)) +
    geom_point(shape = as.factor(celltypes), size = 3) +
    theme_minimal() +
    scale_color_manual(values = standard_colors)+
    labs(title = 'UMAP', color = 'Cell type') + 
    xlab('UMAP 1') +
    ylab('UMAP 2') +
    theme(legend.position = 'right') 
show(umapPlot)
```

```{r updated_umap_plot}

## Reading the data:
umapdata <- data.table::fread(
    "ine/umapdata.tsv",
    header = TRUE)
umapdata <- as.data.frame(umapdata) %>%
    dplyr::mutate(celltypes = gsub(celltypes, pattern = "cells", replacement = " cells"))
## For plotting I am not sure this is neccessary:
rownames(umapdata) <- umapdata$V1 
umapdata$V1 <- NULL
head(umapdata)

umapdata$celltypes <- as.factor(umapdata$celltypes)

## Creatting a "palette" of shapes:
shapes_pal <- c(0:14)

umapPlot_updated <- ggplot2::ggplot(
    umapdata,
    aes(
        x = X1,
        y = X2, 
        group = celltypes)) +
    ggplot2::geom_point(
        aes(
            shape = celltypes,
            color = celltypes),
        size = 3) +
    scale_color_manual(
        values = standard_colors,
        name = "Cell type") +
    scale_shape_manual(
        values = shapes_pal[1:length(unique(umapdata$celltypes))],
        name = "Cell type") +
    labs(
        x = "UMAP 1",
        y = "UMAP 2",
        title = 'UMAP') + 
    theme(legend.position = 'right') 
umapPlot_updated
```


My next question is how to again generate a legend. Now the legend contains the correlation lines but not the shapes. Which for normal people seems fine. But I want to point out the difference. And I prefer to have the legend of the lines on top, so therefore I positioned it on the top right. 

Next to that I was wondering if it would be possible to add a little table with all my R values to the faceted plot. I have seen examples online to put a little table in one of the corners of your plot, similar for every facet of a plot. But I would like to add just one table, next to the facets/panels.

```{r correlation_plot}
allprops <- fread("ine/proportions_for_correlationplot.tsv")
head(allprops)

ssize <- as.factor(allprops$sample_size)
ssize <- factor(ssize, levels = c("5000", "10000", "15000"))

proportions_plot <- ggplot(allprops, aes(x = true_proportion, y = predicted_proportion, col = ssize, shape = ssize)) + 
    geom_point(alpha = 0.5)+
    ggpmisc::stat_poly_line(formula = y ~ x, se = FALSE)+ 
    facet_wrap(~celltype)+
    scale_x_continuous(breaks = seq(0, 0.9, by = 0.2))+
    scale_y_continuous(breaks = seq(0, 0.9, by = 0.2))+
    scale_shape_manual(name = "Cell number input \nfor signature matrix",
                        values = c(16, 17, 18)) + 
    scale_colour_manual(name="Cell number input \nfor signature matrix",
                        values=c("#E41A1C", "#377EB8", "#4DAF4A"))+
    theme(legend.justification = "top", 
        axis.title.x = element_text(vjust = -0.8), 
        axis.title.y = element_text(vjust = 1))+
    xlab("True proportion")+
    ylab("Predicted proportion")
show(proportions_plot)
```